---
title: "Laboratory16"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    math: katex
    df_print: paged
    toc: yes
    toc_depth: 4
             
vignette: >
  %\VignetteIndexEntry{Laboratory16}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "ws#>",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

# directory to Lab
dirdl <- system.file("Lab16",package = "Intro2R")

# create rmd link

library(Intro2R)


```


# Introduction

This lab is about **correlation**,  **adjusted R squared** and **prediction**.

We will be looking at the `Pearson moment correlation coefficient` defined as

$$ r = \frac{SS_{xy}}{\sqrt{SS_{xx}SS_{yy}}} $$
this measures the amount of linear relationship between two variables x, and y.

This is related to another very important statistic, the `coefficient of determination` otherwise known as `multiple r squared`. This is defined as 

$$R^2 = \frac{MSS}{TSS}$$
This is quantifying the fraction of the **TSS** (Total sum of squares = variability of the response = $SS_{yy}$) explained by the model.
We sometimes express this as a percentage, example, if $R^2=0.6456$ then we might say:

>
  64.56% of the variabilty in the response is explained by the model through x
  
  
  For SLR we can derive
  
  $$R^2 = r^2$$

When we wish to predict a new response $y$ given a particular x value $x_p$ or if we wish to predict a mean response $E(Y)$ given an x value $x_p$ we call this **prediction**. There are many functions in R that will help you do this. The main one is `predict()`.

# The data

The data has been packaged in the current library.

```{r}
data("liquid")
head(liquid)
```
## Plot the data

In order to examine the trend in the data we will use a loess smoother:

```{r}
library(ggplot2)
g <- ggplot(liquid, aes(x = TIME, y = MASS)) + geom_point()
g1 <- g + geom_smooth(method = "loess", formula = y~x)
g1
```


The plot has a curved trend. It is likely we should not conclude with a straight line but we could start with this and then make improvements.

## Straight line, quadratic or exponential?

A SLR line would be easy to fit under either paradigm. Lets make a raft of models and name the objects `ylm1, ylm2, ...`
and use adjusted $R^2$ to choose between them.

```{r}
ylm1 <- lm(MASS ~ TIME, data = liquid)
ylm2 <- lm(MASS ~ TIME + I(TIME^2), data = liquid)
ylm3 <- lm(log(MASS+1) ~ TIME + I(TIME^2), data=liquid)
s1 <- summary(ylm1)
s2 <- summary(ylm2)
s3 <- summary(ylm3)
Rasq <- list(Rasq1 = s1$adj.r.squared,
               Rasq2 = s2$adj.r.squared,
               Rasq3 = s3$adj.r.squared)
Rasq

```

The second model is the winner (Largest $R_a^2$) but not by much.

We will use a quadratic to model the relationship.

```{r}
sm <- summary(ylm2)
cf = coef(ylm2)
cf
```


The model is now 

$$\widehat{MASS} = `r cf[1]` + `r cf[2]`*TIME + `r cf[3]`*TIME^2 $$

We can now plot the fitted model

```{r}
g2 <- g + geom_smooth(method = "lm", formula = y~x+I(x^2))
g2 <- g2 + ggtitle("A quadratic model")
g2
```


### Criticism

The model works well within most of the domain of `TIME` but at the extreme end the estimating trend turns up  -- this is a property of the quadratic (coefficient of $TIME^2$ is positive) and would not be of much concern for estimation and prediction within most of the range of TIME. It does give interpretational problems however, at extreme TIME values -- it gives the appearance that MASS would begin to increase with higher TIME.

## Using the exponential model

The model I would like us to look at is the non linear exponential model (you can see that the model is non linear in $\beta_1$ and $\beta_2$)

$$E(MASS|TIME) = \beta_0 e^{\beta_1 TIME + \beta_2 TIME^2}$$
We will use the `nls` function in the `STATS` base package.

```{r}
summary(liquid)
nls1 <- nls(formula = MASS ~ b0*exp(b1*TIME + b2*TIME^2 ),
            data = liquid,
            start = list(b0 = 7, b1 = -0.05,b2  = - 0.0005 ),
            trace = TRUE
            )
summary(nls1)
```

## Plot the exponential model

We will now plot the points and use the estimates created above using Base R.

```{r message=TRUE, warning=TRUE}
cf <- coef(nls1)

mynlsfun <- function(x) cf[1]*exp(cf[2]*x + cf[3]*x^2)

plot(MASS ~ TIME, data = liquid, main = "Non linear Least Squares")

curve(mynlsfun, lwd = 3, col = "Blue", add = TRUE)
text(30,4, expression(hat(E)(MASS)==hat(beta)[0]*exp(hat(beta)[1]*TIME + hat(beta)[2]*TIME)))
```


# Finally 

With this brief introduction you may now start on the lab

`r rmdfiles("Lab16","Intro2R")`
