---
title: "Laboratory2"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    math: katex
    df_print: paged
    toc: yes
             
vignette: >
  %\VignetteIndexEntry{Laboratory2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "ws#>"
)

# directory to Lab
dirdl <- system.file("Lab2",package = "Intro2R")

# create rmd link

library(Intro2R)
dir(dirdl)

```



# Introduction

`r rmdfile("Lab 2","MATH4753Laboratory2.docx","Lab2")` will cement a number of R skills and theory related to the following:

  1.	The empirical rule
  2.	The Chebyshev rule
  3.	Transformation of data to z values
  4.	Location of outliers using z values
  

# R skills

For the empirical rule you will need to be able to calculate means and standard deviations.

You can make z values in a number of different ways.

## Z, Method 1

Suppose we wish to make Z values from the DDT variable in the `ddt` data frame.

```{r}
d <- ddt$DDT
z <- (d-mean(d))/sd(d)
head(z)
```

## Z, Method 2

We can use a built in function called `scale()`

```{r}
zmat<-scale(ddt$DDT) # scale makes a matrix of z values
z<-zmat[,1] # take the column to form a vector
head(z)
```

## Verify Chebyshev and Empirical rules

To show how many data values lie within `k` standard deviations of the mean you will need to count the number of values in an interval

### Making intervals

```{r}
mn <- mean(ddt$DDT)
sdd <- sd(ddt$DDT)
mp<-c(-1,1)
k<-2

mn+mp*k*sdd

```

### Values within k standard deviations of the mean

We can subset the data frame with `abs(z)<2` and then pull off the DDT column.
The function `length` adds up the number of values in the vector

```{r}
ddl2<-ddt[abs(z)<2,"DDT"]
length(ddl2)/length(ddt$DDT)*100
```
98.6% of the DDT lies within 2 standard deviations of the mean.

# Dotplot in ggplot

The dotplot is a simple device, we can set bins of a certain size (I will use 1/5 sd(LENGTH)) to break the continuous data into discrete levels.

In addition we will use the `cut()` function to create labels corresponding to regions of LENGTH that are distant from the mean by integral standard deviations.

See the code below:

```{r}
library(ggplot2)
library(dplyr)
mn <- mean(ddt$LENGTH)
sdd <- sd(ddt$LENGTH)
ddt <- ddt %>% mutate( Lcut = cut(LENGTH, c(min(LENGTH)-1, 
                                            seq(mn-3*sdd, mn+3*sdd, by = sdd), max(LENGTH)),
                                  labels = c("Outlier","Possible Out","Main","Main", "Main","Main", "Possible Out", "Outlier")))
g <- ggplot(ddt, aes(x = LENGTH)) + geom_dotplot(aes(fill = Lcut),binwidth = 1/5*sdd)
g <- g + geom_density(aes(y = ..count..))
g <- g + labs(title = "LENGTH data categorized by outlier status using z")
g
```


# Instructions

You will need the following files

```{r}
dir(dirdl)
```

Make sure you place them all in Lab2.

`r rmdfile("EPAGAS", "EPAGAS.XLS","Lab2")`

`r rmdfile("Lab2.R", "Lab2.R", "Lab2")`

`r rmdfile("Lab document","MATH4753Laboratory2.docx","Lab2")`


